---
title: "Weather events versus health and economy"
author: "Andrew Crisp"
date: "Monday, August 17, 2015"
output: html_document
---

# Synopsis


Given the NOAA Storm Data, demonstrate:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

Knowing the answers to these questions will aid in preparations.  Tornadoes present the most injuries and fatalities.  Tornadoes mixed with Thunderstorm Wind and Hail cause the most economic damage.

# Data Processing

## Let's load the data
```{r cache = TRUE}
if(!file.exists("repdata-data-StormData.csv.bz2"))
{
  download.file(url="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                destfile = "repdata-data-StormData.csv.bz2")
}
StormData <- read.csv("repdata-data-StormData.csv.bz2")
 
```

## Question 1

### Across the United States, which types of events are most harmful with respect to population health?

Let's combine Fatalities and Injuries into a single count, Casualties.  This allows for easy sorting by maximum value.  

```{r}
options(scipen = 9)
CasualtyData <- StormData[,c("EVTYPE","FATALITIES","INJURIES")]
CasualtyData$CASUALTIES <- (CasualtyData$FATALITIES + CasualtyData$INJURIES)

byCasualties <- aggregate(list(Casualties = CasualtyData$CASUALTIES,
                               Injuries = CasualtyData$INJURIES,
                               Fatalities = CasualtyData$FATALITIES),
                          list(EventType = CasualtyData$EVTYPE),
                          FUN = sum)
byCasualties <- byCasualties[order(byCasualties$Casualties,decreasing = TRUE),]

```

## Question 2

### Across the United States, which types of events have the greatest economic consequences?

The provided data had many incorrectly entered exponent values.  This will convert them to numeric powers of ten.  The odd values "+", "-", and "?" are converted to 0.  Essentially, this completely discounts their values.  But, there is such a small number of events that this is nearly unnoticed.

After calculating the dollar cost of Property and Crop damage, combine them into a single field for sorting of maximum damage.

```{r}
economicData <- StormData[,c("EVTYPE","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]
economicData$PROPDMGEXP <- gsub(pattern = "+", replacement = "0", economicData$PROPDMGEXP, fixed = TRUE)
economicData$PROPDMGEXP <- gsub(pattern = "-", replacement = "0", economicData$PROPDMGEXP, fixed = TRUE)
economicData$PROPDMGEXP <- gsub(pattern = "?", replacement = "0", economicData$PROPDMGEXP, fixed = TRUE)
economicData$PROPDMGEXP <- gsub(pattern = "m", replacement = "M", economicData$PROPDMGEXP, fixed = TRUE)
economicData$PROPDMGEXP <- gsub(pattern = "k", replacement = "K", economicData$PROPDMGEXP, fixed = TRUE)
economicData$PROPDMGEXP <- gsub(pattern = "h", replacement = "H", economicData$PROPDMGEXP, fixed = TRUE)
economicData$PROPDMGEXP <- gsub(pattern = "M", replacement = "6", economicData$PROPDMGEXP, fixed = TRUE)
economicData$PROPDMGEXP <- gsub(pattern = "K", replacement = "3", economicData$PROPDMGEXP, fixed = TRUE)
economicData$PROPDMGEXP <- gsub(pattern = "H", replacement = "2", economicData$PROPDMGEXP, fixed = TRUE)
economicData$PROPDMGEXP <- gsub(pattern = "B", replacement = "9", economicData$PROPDMGEXP, fixed = TRUE)
economicData$PropertyDamage <- (economicData$PROPDMG * (10^as.numeric(economicData$PROPDMGEXP)))

economicData$CROPDMGEXP <- gsub(pattern = "?", replacement = "0", economicData$CROPDMGEXP, fixed = TRUE)
economicData$CROPDMGEXP <- gsub(pattern = "k", replacement = "K", economicData$CROPDMGEXP, fixed = TRUE)
economicData$CROPDMGEXP <- gsub(pattern = "m", replacement = "M", economicData$CROPDMGEXP, fixed = TRUE)
economicData$CROPDMGEXP <- gsub(pattern = "B", replacement = "9", economicData$CROPDMGEXP, fixed = TRUE)
economicData$CROPDMGEXP <- gsub(pattern = "K", replacement = "3", economicData$CROPDMGEXP, fixed = TRUE)
economicData$CROPDMGEXP <- gsub(pattern = "M", replacement = "6", economicData$CROPDMGEXP, fixed = TRUE)
economicData$CropDamage <- (economicData$CROPDMG * (10^as.numeric(economicData$CROPDMGEXP)))

economicData$TotalDamage <- (economicData$PropertyDamage + economicData$PropertyDamage)
byDamage <- aggregate(list(TotalDamage = economicData$TotalDamage,
                           PropertyDamage = economicData$PropertyDamage,
                           CropDamage = economicData$CropDamage),
                      list(EventType = economicData$EVTYPE),
                      FUN = sum)
byDamage <- byDamage[order(byDamage$TotalDamage,decreasing = TRUE),]


```

# Results

## Question 1

### Across the United States, which types of events are most harmful with respect to population health?
```{r}
with(byCasualties[1:3,],barplot(height = Casualties,
                                names.arg = EventType,
                                main = "Highest total injuries and fatalities", 
                                cex.names = .7,
                                xlab = "Event Type", 
                                ylab = "Total"))

byCasualties[1:5,]

```


We can see that Tornadoes represent the highest human damage by a significant amount at `r byCasualties[1,2]` incidents.  Excessive heat and Thunderstorm Wind combined still represent a fraction of the damage of Tornadoes, at `r byCasualties[2,2]` and `r byCasualties[3,2]` each.

## Question 2

### Across the United States, which types of events have the greatest economic consequences?

```{r}
with(byDamage[1:3,],barplot(height = TotalDamage,
                           names.arg = EventType,
                           main = "Highest economic cost",
                           cex.names = .7,
                           xlab = "Event Type",
                           ylab = "Total cost"))

byDamage[1:5,]

```

As for economic cost, a combination of Tornadoes, Thunderstorm Wind, and Hail makes the largest source of damage at $`r byDamage[1,2]`.  Wild Fires and Hailstorms place second and third at $`r byDamage[2,2]` and $`r byDamage[3,2]`, respectively.  Again, the primary damage source is more than twice the second and third sources combined.